\section{Introduction}
% ─────────────────────────────────────────────────────────────────

The Transformer \cite{vaswani2017attention} discards recurrence entirely,
relying solely on attention mechanisms to model dependencies between positions.
This section motivates the architecture and outlines the contributions of this
assignment.

\begin{defn}[Scaled Dot-Product Attention]
Given queries $\bm{Q} \in \R^{n \times d_k}$, keys $\bm{K} \in \R^{m \times d_k}$,
and values $\bm{V} \in \R^{m \times d_v}$, attention is defined as
\begin{equation}
    \Attn(\bm{Q}, \bm{K}, \bm{V}) =
        \softmax\!\left(\frac{\bm{Q}\bm{K}^\top}{\sqrt{d_k}}\right)\bm{V}.
    \label{eq:attention}
\end{equation}
\end{defn}

The $\sqrt{d_k}$ scaling prevents the dot products from entering saturation
regions of the softmax when $d_k$ is large.

% ─────────────────────────────────────────────────────────────────
\section{Architecture}
% ─────────────────────────────────────────────────────────────────

\subsection{Multi-Head Attention}

Rather than computing a single attention function, the model projects $\bm{Q}$,
$\bm{K}$, $\bm{V}$ $h$ times with learned linear projections, attends in
parallel, then concatenates:

\begin{eqbox}
\begin{equation}
    \MHA(\bm{Q},\bm{K},\bm{V}) =
        \operatorname{Concat}(\text{head}_1, \ldots, \text{head}_h)\,\bm{W}^O
    \label{eq:mha}
\end{equation}
\end{eqbox}

where $\text{head}_i = \Attn(\bm{Q}\bm{W}_i^Q,\, \bm{K}\bm{W}_i^K,\,
\bm{V}\bm{W}_i^V)$.

\begin{thm}[Expressivity of MHA]
Multi-head attention with $h$ heads and $d_{\text{model}} = h\,d_k$ parameters
is strictly more expressive than single-head attention with the same parameter
budget, as each head can attend to a distinct subspace of the representation.
\end{thm}

\subsection{Position-wise Feed-Forward Network}

Each encoder/decoder layer includes a two-layer FFN applied identically to
each position:
\begin{equation}
    \operatorname{FFN}(\bm{x}) = \max(0,\, \bm{x}\bm{W}_1 + \bm{b}_1)\bm{W}_2 + \bm{b}_2
\end{equation}

\begin{remark}[Dimensionality]
The inner dimension is typically $d_{ff} = 4\,d_{\text{model}}$.
In the base model \cite{vaswani2017attention}, $d_{\text{model}} = 512$
and $d_{ff} = 2048$.
\end{remark}

\subsection{Positional Encoding}

Since attention is permutation-equivariant, position information must be
injected explicitly. Sinusoidal encodings are given by:
\begin{align}
    \text{PE}_{(pos,\,2i)}   &= \sin\!\left(pos / 10000^{2i/d}\right) \\
    \text{PE}_{(pos,\,2i+1)} &= \cos\!\left(pos / 10000^{2i/d}\right)
\end{align}

\begin{important}[Learned vs.\ Sinusoidal]
Learned positional embeddings and sinusoidal encodings achieve comparable
performance; however, sinusoidal encodings generalise to sequence lengths
unseen during training.
\end{important}

% ─────────────────────────────────────────────────────────────────
\section{Implementation}
% ─────────────────────────────────────────────────────────────────

\subsection{Attention Module}

The core attention function in PyTorch:

\begin{codebox}[Python — Scaled Dot-Product Attention]
import torch
import torch.nn.functional as F
import math

def scaled_dot_product_attention(Q, K, V, mask=None):
    """
    Q, K, V : (batch, heads, seq, d_k)
    """
    d_k = Q.size(-1)
    scores = torch.matmul(Q, K.transpose(-2, -1)) / math.sqrt(d_k)
    if mask is not None:
        scores = scores.masked_fill(mask == 0, float('-inf'))
    weights = F.softmax(scores, dim=-1)
    return torch.matmul(weights, V), weights
\end{codebox}

\subsection{Layer Normalisation}

Residual connections are followed by layer normalisation:
\begin{equation}
    \bm{y} = \LayerNorm\!\left(\bm{x} + \operatorname{Sublayer}(\bm{x})\right)
\end{equation}

\begin{remark}[Pre-LN vs.\ Post-LN]
Modern implementations (GPT, LLaMA) typically apply normalisation
\emph{before} the sublayer (\emph{Pre-LN}), which improves gradient flow and
stabilises training without a learning rate warm-up.
\end{remark}

% ─────────────────────────────────────────────────────────────────
\section{Training Dynamics}
% ─────────────────────────────────────────────────────────────────

\subsection{Learning Rate Schedule}

The original Transformer uses a warm-up schedule:
\begin{equation}
    \alpha = d_{\text{model}}^{-0.5}
              \cdot \min\!\left(\mathit{step}^{-0.5},\;
              \mathit{step} \cdot \mathit{warmup}^{-1.5}\right)
\end{equation}

\begin{important}[Warm-up Critical]
Without the linear warm-up phase the model often diverges in early training.
A typical value is $\mathit{warmup} = 4000$ steps.
\end{important}

\subsection{Regularisation}

Three regularisation strategies are employed:

\begin{enumerate}[leftmargin=*, itemsep=2pt, topsep=3pt]
    \item \textbf{Dropout} applied to attention weights and sublayer outputs
          ($p_{\text{drop}} = 0.1$).
    \item \textbf{Label smoothing} with $\varepsilon_{ls} = 0.1$, which hurts
          perplexity but improves BLEU.
    \item \textbf{Weight decay} ($\lambda = 10^{-4}$) on all non-bias parameters.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────
\section{Experimental Results}
% ─────────────────────────────────────────────────────────────────

\subsection{Ablation: Number of Heads}

\begin{table}[H]
\centering
\small
\setlength{\tabcolsep}{6pt}
\begin{tabular}{lccc}
\toprule
\rowcolor{altbgraised}
\textbf{Heads} ($h$) & \textbf{Val.\ Loss} & \textbf{BLEU} & \textbf{Params} \\
\midrule
1  & 3.42 & 18.1 & 37.4M \\
\rowcolor{altrowave}
4  & 2.91 & 23.7 & 37.4M \\
8  & 2.78 & 25.3 & 37.4M \\
\rowcolor{altrowave}
16 & 2.81 & 24.9 & 37.4M \\
\bottomrule
\end{tabular}
\caption{Effect of the number of attention heads (base model, $d_{\text{model}}=512$).}
\label{tab:heads}
\end{table}

Eight heads yields the best BLEU score; increasing to 16 slightly degrades
performance, consistent with findings in \cite{voita2019analyzing}.

% ─────────────────────────────────────────────────────────────────
\section{Conclusion}
% ─────────────────────────────────────────────────────────────────

We have implemented and analysed the Transformer architecture from scratch,
covering attention, positional encodings, layer normalisation, and training
dynamics. Ablation results confirm the importance of multi-head attention and
proper learning rate scheduling. Future work will explore relative positional
encodings and sparse attention variants.

% ─────────────────────────────────────────────────────────────────
%  REFERENCES  (replace with .bib file as needed)
% ─────────────────────────────────────────────────────────────────
\begin{thebibliography}{9}
\small
\bibitem{vaswani2017attention}
A.\ Vaswani et al., ``Attention is all you need,''
\textit{NeurIPS}, 2017.

\bibitem{voita2019analyzing}
E.\ Voita et al., ``Analyzing multi-head self-attention: Specialized heads
do the heavy lifting, the rest can be pruned,''
\textit{ACL}, 2019.
\end{thebibliography}
